Detection and replication of epistasis influencing transcription in humans

### 2013-06-20 docs/manuscript/methods.tex
updated manuscript and methods

We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013, Rietvald2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions. We subsequently refer to these 731 SNPs as the IndexSNPs A recent paper shows that H3K4me3 chromatin marks are the most cell-type-specific marks in terms of co-localization with previously published GWAS loci \cite{Trynka2013}. The idea behind the method is that variants related to a particular phenotype may affect cell-type-specific gene expression by changing regulatory elements in cell types relevant to that phenotype. Hence, overlap between associated variants and chromatin marks should occur preferentially in the relevant cell type(s). Our analysis tested for enrichment of these chromatin marks in 34 different cell types from 4 different broad tissue types.

For each IndexSNP, a locus region was defined, bounded on either side of the IndexSNP by the most distant SNP within 250 kb that was in LD (r2 > 0.8). For each SNP in a locus region regulatory activity scores were calculated as the height of the nearest H3K4me3 mark divided by distance from the SNP to the H3K4me3 mark. The SNP with the highest score within each locus region in a given tissue was designated the BestSNP, which served as the score representing that locus.

@@

@@

\subsection{SNP colocalisation with genic features}

We tested for enrichment of genic features for the 731 SNPs that comprise the 501 epistatic interactions. For each of the 731 SNPs we calculated LD with all neighbouring SNPs within a radius of 1kb and kept all neighbouring SNPs with LD $r^2 > 0.8$. We then cross-referenced the remaining regional SNPs with the annotated reference genome (version HG19) querying whether the regional SNPs fell in intergenic, promotor, intronic, exonic, splice sites, 5' UTR or 3' UTR positions. Therefore a particular query SNP might have regional SNPs in multiple genic features.

We then performed the whole querying process for each of the 528,509 SNPs present in the SNP chip used in the scan, and used the results from this second analysis to establish the expected proportion of SNPs for each genic feature.

>>>>

We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions. We subsequently refer to these 731 SNPs as the IndexSNPs A recent paper shows that H3K4me3 chromatin marks are the most cell-type-specific marks in terms of co-localization with previously published GWAS loci \cite{Trynka2013}. The idea behind the method is that variants related to a particular phenotype may affect cell-type-specific gene expression by changing regulatory elements in cell types relevant to that phenotype. Hence, overlap between associated variants and chromatin marks should occur preferentially in the relevant cell type(s). Our analysis tested for enrichment of these chromatin marks in 34 different cell types from 4 different broad tissue types.

For each IndexSNP, a locus region was defined, bounded on either side of the IndexSNP by the most distant SNP within 250 kb that was in LD ($r^{2} > 0.8$). For each SNP in a locus region regulatory activity scores were calculated as the height of the nearest H3K4me3 mark divided by distance from the SNP to the H3K4me3 mark. The SNP with the highest score within each locus region in a given tissue was designated the BestSNP, which served as the score representing that locus.

@@

\subsection{SNP colocalisation with genomic features}
\label{sub:snp_colocalisation_with_genomic_features}

We tested for enrichment of genomic features for the 687 IndexSNPs that comprise the 434 epistatic interactions with data present in discovery and replication datasets. For each of the 687 IndexSNPs we calculated LD with all regional SNPs within a radius of 0.5Mb and kept all regional SNPs with LD $r^{2} > 0.8$. We then cross-referenced the remaining regional SNPs with the annotated reference genome (version HG19, GENCODE annotation version 16 \cite{Harrow2012}) querying whether the regional SNPs fell in intergenic, promotor, intronic, exonic, splice sites, 5' UTR or 3' UTR positions. Therefore a particular IndexSNP might cover multiple genomic features.

We then performed the whole querying process for each of the 528,509 SNPs present in the SNP chip used in the scan, and used the results from this second analysis to establish a null distribution for the expected proportion of SNPs for each genomic feature. We calculated $p$-values for enrichment of each of the seven genomic features independently, and for \emph{cis}- and \emph{trans}-SNPs separately, using a binomial test. For each genomic feature we used the expected proportion of SNPs as the expected probability of ``success'' ($p$). Here, a success is defined as an IndexSNP residing in a region that includes the genomic feature. The observed number of successes for each IndexSNP ($k$) out of the total count of IndexSNPs ($n$) was then modelled as $\Pr (X = k) = {n \choose k} p^{k}(1-p)^{n-k}$.

@@

To test for enrichment of transcription factor binding sites (TFBS) we followed a procedure similar to that described in Section \ref{sub:snp_colocalisation_with_genomic_features}. For each of the 687 IndexSNPs we extracted regional SNPs as previously described. We then used the PWMEnrich package in Bioconductor (http://www.bioconductor.org/packages/2.12/bioc/html/PWMEnrich.html) to identify which TFBSs each of the regional SNPs for one IndexSNP falls in (within a radius of 250bp). Thus, the number of occurrences of a particular TFBS was counted for each IndexSNP. We used the ``Threshold-free affinity'' method for identifying TFBSs \cite{Stormo2000}.

We constructed a null distribution of expected TFBS occurrences based on the same null hypothesis as described in Section \ref{sub:snp_colocalisation_with_genomic_features} - the probability of an IndexSNP covering a particular TFBS is identical to any of the 528,509 SNPs in the discovery SNP chip. To do this, we performed the same procedure for each SNP in the discovery SNP chip as was performed for each IndexSNP to obtain an expected probability of covering a particular TFBS. We then tested the IndexSNPs for enrichment of each TFBS independently, and for \emph{cis}- and \emph{trans}-SNPs separately. $p$-values were obtained using Z-scores, calculated by using a normal approximation to the sum of binomial random variables representing motif hits along the sequence \cite{HoSui2005}.

### 2013-06-20 docs/manuscript/nature_letter.tex
updated manuscript and methods

\title{Detection and replication of epistasis influencing transcription in humans}
\date{}
\author{Gibran Hemani, Joseph E Powell, Konstantin Shakhbazov, \\
Greg Gibson, Grant W Montgomery, Peter M Visscher}

@@

Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting. We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP. For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect. We observed that nine of the 14 masking SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$, Supplementary Figure \ref{fig:snplocations}). Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).

@@

We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.

@@

\subsection{Methods Summary}
We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7,339 probes representing 6,158 RefSeq genes. Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis. We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}. Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone. All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs. The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$). Of these, 434 passed filtering in both replication datasets. A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

>>>>

\title{Detection and replication of epistasis influencing transcription in humans}
\date{}
\author{Gibran Hemani, Joseph E Powell, Konstantin Shakhbazov, \\
Harm-Jan Westra, Anjali Henders, Allan F. McRae, \\
Andres Metspalu, Greg Gibson, Tonu Esko, Lude Franke,  \\
Nick Martin, Grant W Montgomery, Peter M Visscher}

@@

Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting. We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP. For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect. Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}). We see similar epistatic networks involving multiple \emph{trans}-acting SNPs for

@@

We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}, and that `transcriptional factories' can organise across distant chromosomal regions \cite{Osborne2004}. We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96,139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.

@@

\subsection{Methods Summary}
We searched for pairwise epistasis exhaustively in the BSGS discovery dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528,509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7,339 probes representing 6,158 RefSeq genes. Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis. We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}. Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone. All SNP pairs with LD $r^2 > 0.1$ and $D'^{2} > 0.1$ were removed to minimise the possibility of haplotype effects. All SNP pairs were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs. The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$). Of these, 434 passed filtering in both replication datasets. A meta analysis on the interaction $p$-values from each replication dataset was performed to provide an overall replication statistic for each putative interaction.

\subsection{Acknowledgements}

We are grateful to the volunteers for their generous participation in these studies, and thank Bill Hill, Chris Haley and Lars Ronnegard for helpful discussions and comments.

The University of Queensland group is supported by the Australian Research Council (DP130102666, DE130100691), the Australian National Health and Medical Research Council (APP1011506, APP1047956, APP1048853) and the National Institutes of Health (GM099568, GM075091).

The QIMR researchers acknowledge funding from the Australian National Health and Medical Research Council (grants 241944, 389875, 389891, 389892, 389938, 442915, 442981, 496739, 496688 and 552485), the and the National Institutes of Health (grants AA07535, AA10248, AA014041, AA13320, AA13321, AA13326 and DA12854).

This work could not have been completed without access to high performance GPGPU compute clusters. We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au). We also acknowledge Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.

The CHDWB study funding support from the Georgia Institute of Technology Research Foundation. The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript

The Fehrmann study was supported by grants from the Celiac Disease Consortium (an innovative cluster approved by the Netherlands Genomics Initiative and partly funded by the Dutch Government (grant BSIK03009), the Netherlands Organization for Scientific Research (NWO-VICI grant 918.66.620, NWO-VENI grant 916.10.135 to L.F.), the Dutch Digestive Disease Foundation (MLDS WO11-30), and a Horizon Breakthrough grant from the Netherlands Genomics Initiative (grant 92519031 to L.F.). This project was supported by the Prinses Beatrix Fonds, VSB fonds, H. Kersten and M. Kersten (Kersten Foundation), The Netherlands ALS Foundation, and J.R. van Dijk and the Adessium Foundation. The research leading to these results has received funding from the European Community’s Health Seventh Framework Programme (FP7/2007-2013) under grant agreement 259867.

The EGCUT study received targeted financing from Estonian Government SF0180142s08, Center of Excellence in Genomics (EXCEGEN) and University of Tartu (SP1GVARENG). We acknowledge EGCUT technical personnel, especially Mr V. Soo and S. Smit. Data analyzes were carried out in part in the High Performance Computing Center of University of Tartu.
